---
title: "Effects of Past COVID-19 Exposure on Cognitive Processing: An EZ Bayesian Hierarchical Drift Diffusion Model Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# Abstract

The COVID-19 pandemic has raised concerns about lasting cognitive effects following infection. This study investigates whether past COVID-19 exposure is associated with changes in cognitive processing using an EZ Bayesian hierarchical drift diffusion model (EZBHDDM). We analyzed data from 68 participants who completed a change-detection task with a $2 \times 2 \times 2 \times 2$ factorial design. The hierarchical Bayesian model simultaneously estimated the effects of COVID-19 exposure on three key cognitive parameters: drift rate (processing efficiency), boundary separation (response caution), and non-decision time (encoding and motor processes).

**Keywords:** COVID-19, cognitive deficits, drift diffusion model, EZ-diffusion, hierarchical Bayesian modeling, response time

---

# Introduction

The COVID-19 pandemic, caused by the SARS-CoV-2 virus, has affected millions worldwide. Beyond acute respiratory symptoms, accumulating evidence suggests that COVID-19 can have lasting effects on cognitive function. Hampshire, Trender, Chamberlain, and colleagues (2021) demonstrated in a large-scale study ($N = 84{,}285$) that individuals who had recovered from COVID-19, including those with mild cases not requiring hospitalization, exhibited significant cognitive deficits across multiple domains compared to matched controls. These findings align with reports of "long COVID" or post-acute sequelae of SARS-CoV-2 infection (PASC), which frequently include cognitive symptoms such as "brain fog," attention difficulties, and slowed processing (Hampshire et al., 2021).

While the existence of COVID-related cognitive deficits is increasingly well-documented, the underlying mechanisms remain unclear. Traditional cognitive assessments provide aggregate measures of performance (e.g., mean accuracy or response time) but do not distinguish between different cognitive processes that may be affected. The drift diffusion model (DDM; Ratcliff, 1978) offers a formal framework for decomposing decision-making performance into psychologically interpretable parameters. The model posits that decisions arise from the accumulation of evidence toward one of two response boundaries, with three key parameters: drift rate ($v$), reflecting the quality and speed of evidence accumulation; boundary separation ($a$), reflecting response caution or the speed-accuracy tradeoff; and non-decision time ($T_{er}$), reflecting encoding and motor processes.

Wagenmakers, van der Maas, and Grasman (2007) introduced the EZ-diffusion model, which provides closed-form estimators for DDM parameters from summary statistics (mean accuracy, mean response time, and RT variance). Building on this foundation, Ch\'avez De la Pe\~na and Vandekerckhove (2025) developed the EZ Bayesian hierarchical drift diffusion model (EZBHDDM), which integrates the EZ-diffusion equations into a Bayesian hierarchical framework. This approach offers several advantages: it properly accounts for uncertainty at all levels, enables principled model comparison, and allows for regression structures that relate DDM parameters to covariates of interest.

The present study applies the EZBHDDM to investigate the effects of past COVID-19 exposure on cognitive processing in a change-detection task. We test whether COVID-19 exposure is associated with changes in drift rate (suggesting impaired evidence accumulation), boundary separation (suggesting altered response caution), or non-decision time (suggesting slower encoding or motor processes). By applying a hierarchical Bayesian approach, we account for individual differences while leveraging the experimental design's factorial structure.

---

# Method

## Participants

The sample comprised 68 participants (after excluding one participant with insufficient trial data). Participants provided demographic information including age, sex, race, ethnicity, and education level, as well as COVID-19 exposure history.

## Experimental Task

Participants completed a change-detection task with a $2 \times 2 \times 2 \times 2$ factorial design manipulating four binary factors:

1. **Study Time (ST)**: Short (500ms) vs. Long (2000ms) stimulus presentation
2. **Choice Urgency (CU)**: High urgency (500ms deadline) vs. Low urgency (3000ms deadline)
3. **Probability of Change (PC)**: Equal probability (50/50) vs. Unequal probability (20/80)
4. **Probe Type (PT)**: Whole display vs. Single probe

These manipulations were designed to target different components of the decision process: Study Time affects evidence quality (predicted to influence drift rate), Choice Urgency affects speed-accuracy tradeoff (predicted to influence boundary separation), and Probability of Change affects prior bias. Each participant completed up to 960 trials across the 16 experimental conditions.

## Data Processing

Trial-level data were preprocessed as follows. Trials with response times below 200ms or above 7000ms were excluded as anticipatory or outlier responses. For each participant and condition, we computed the mean response time (in seconds), response time variance, total number of correct responses, and number of trials. Demographic data (including COVID status) were merged with the cognitive task summaries. Continuous covariates (age, education) were standardized for model estimation.

## EZ Bayesian Hierarchical Drift Diffusion Model

### Model Structure

We implemented an integrated EZBHDDM following Ch\'avez De la Pe\~na and Vandekerckhove (2025). The model consists of three interconnected components:

**1. Meta-regression structure for latent DDM parameters.**
For each observation $p$ from participant $i$:

\begin{align*}
v_{p} &= \mu_v + \delta_{v,i} + \beta_{v,\text{COVID}} \cdot \text{COVID}_i + \beta_{v,\text{ST}} \cdot \text{ST}_p + \beta_{v,\text{PC}} \cdot \text{PC}_p + \beta_{v,\text{PT}} \cdot \text{PT}_p + \text{demographics} \\
a_{p} &= \mu_a + \delta_{a,i} + \beta_{a,\text{COVID}} \cdot \text{COVID}_i + \beta_{a,\text{CU}} \cdot \text{CU}_p + \beta_{a,\text{PT}} \cdot \text{PT}_p + \text{demographics} \\
T_{er,p} &= \mu_{T_{er}} + \delta_{T_{er},i} + \beta_{T_{er},\text{COVID}} \cdot \text{COVID}_i + \beta_{T_{er},\text{PT}} \cdot \text{PT}_p + \text{demographics}
\end{align*}

where $\delta_{\cdot,i}$ are participant-specific random effects.

**2. Forward EZ-DDM equations** (deterministic):

\begin{align*}
P_c &= \frac{1}{1 + e^{-av}},  \quad
\text{MRT} = \frac{a}{2v} \cdot \frac{1 - e^{-av}}{1 + e^{-av}} + T_{er}, \quad
\text{PRT} = \frac{2v^3}{a} \cdot \frac{(e^{-av} + 1)^2}{1 - 2av \cdot e^{-av} - e^{-2av}}
\end{align*}

**3. Sampling distributions** for observed statistics:

\begin{align*}
\dot{T} &\sim \text{Binomial}(P_c, N), \quad
\dot{V} \sim \text{Normal}\!\left(\frac{1}{\text{PRT}},\; \frac{N-1}{2 \cdot \text{PRT}^2}\right), \quad
\dot{M} \sim \text{Normal}(\text{MRT},\; \text{PRT} \cdot N)
\end{align*}

### Estimation

The model was implemented in JAGS (Plummer, 2003) via the R2jags interface. We ran 4 MCMC chains with 4000 iterations each, discarding the first 2000 as burn-in. Convergence was assessed via $\hat{R}$ statistics.

---

# Analysis

JAGS results are pre-computed by `run_jags.R` and cached in
`../cache/jags_samples.rds`.  To regenerate, run `Rscript run_jags.R`
from the `src/` directory.

```{r load-cache}
library(ggplot2)
library(knitr)

cache_file <- "../cache/jags_samples.rds"
stopifnot(file.exists(cache_file))
cached <- readRDS(cache_file)

samples       <- cached$samples
data_for_jags <- cached$data_for_jags

cat("Cache timestamp:", format(cached$timestamp), "\n")
cat("Participants:", length(unique(data_for_jags$participant_id)), "\n")
cat("Observations:", nrow(data_for_jags), "\n")
cat("COVID+:", sum(data_for_jags$covid == 1),
    " COVID-:", sum(data_for_jags$covid == 0), "\n")
```

## Convergence Diagnostics

```{r convergence}
jags_summary <- samples$BUGSoutput$summary

key_params <- c("drift_mean", "bound_mean", "nondt_mean",
                "beta_covid_drift", "beta_covid_bound", "beta_covid_nondt",
                "beta_ST_drift", "beta_CU_bound", "beta_PC_drift")

kable(round(jags_summary[key_params, c("mean", "sd", "2.5%", "97.5%", "Rhat", "n.eff")], 3),
      caption = "Posterior summaries and convergence diagnostics for key parameters.")
```

## COVID Effects on DDM Parameters

```{r covid-effects}
covid_drift <- samples$BUGSoutput$sims.list$beta_covid_drift
covid_bound <- samples$BUGSoutput$sims.list$beta_covid_bound
covid_nondt <- samples$BUGSoutput$sims.list$beta_covid_nondt

covid_summary <- data.frame(
  Parameter = c("Drift Rate (v)", "Boundary (a)", "Non-decision Time (Ter)"),
  Mean = c(mean(covid_drift), mean(covid_bound), mean(covid_nondt)),
  SD = c(sd(covid_drift), sd(covid_bound), sd(covid_nondt)),
  `Lower 95` = c(quantile(covid_drift, .025), quantile(covid_bound, .025), quantile(covid_nondt, .025)),
  `Upper 95` = c(quantile(covid_drift, .975), quantile(covid_bound, .975), quantile(covid_nondt, .975)),
  `P(>0)` = c(mean(covid_drift>0), mean(covid_bound>0), mean(covid_nondt>0)),
  check.names = FALSE
)
kable(covid_summary, digits = 3, caption = "Posterior estimates of COVID effects on DDM parameters.")
```

```{r plot-covid, fig.width=10, fig.height=4}
plot_df <- data.frame(
  value = c(covid_drift, covid_bound, covid_nondt),
  param = rep(c("Drift Rate (v)", "Boundary (a)", "Non-decision Time (Ter)"),
              each = length(covid_drift))
)
ggplot(plot_df, aes(x = value, fill = param)) +
  geom_density(alpha = .6) +
  geom_vline(xintercept = 0, lty = 2) +
  facet_wrap(~param, scales = "free_x") +
  labs(title = "Posterior distributions of COVID effects on DDM parameters",
       x = "Effect size", y = "Density") +
  theme_minimal() + theme(legend.position = "none")
```

## Experimental Manipulation Effects

```{r exp-effects}
exp_effects <- data.frame(
  Effect = c("Study Time -> Drift", "Choice Urgency -> Boundary",
             "Prob Change -> Drift", "Probe Type -> Drift",
             "Probe Type -> Boundary", "Probe Type -> Nondt"),
  Mean = c(mean(samples$BUGSoutput$sims.list$beta_ST_drift),
           mean(samples$BUGSoutput$sims.list$beta_CU_bound),
           mean(samples$BUGSoutput$sims.list$beta_PC_drift),
           mean(samples$BUGSoutput$sims.list$beta_PT_drift),
           mean(samples$BUGSoutput$sims.list$beta_PT_bound),
           mean(samples$BUGSoutput$sims.list$beta_PT_nondt)),
  Lower = c(quantile(samples$BUGSoutput$sims.list$beta_ST_drift, .025),
            quantile(samples$BUGSoutput$sims.list$beta_CU_bound, .025),
            quantile(samples$BUGSoutput$sims.list$beta_PC_drift, .025),
            quantile(samples$BUGSoutput$sims.list$beta_PT_drift, .025),
            quantile(samples$BUGSoutput$sims.list$beta_PT_bound, .025),
            quantile(samples$BUGSoutput$sims.list$beta_PT_nondt, .025)),
  Upper = c(quantile(samples$BUGSoutput$sims.list$beta_ST_drift, .975),
            quantile(samples$BUGSoutput$sims.list$beta_CU_bound, .975),
            quantile(samples$BUGSoutput$sims.list$beta_PC_drift, .975),
            quantile(samples$BUGSoutput$sims.list$beta_PT_drift, .975),
            quantile(samples$BUGSoutput$sims.list$beta_PT_bound, .975),
            quantile(samples$BUGSoutput$sims.list$beta_PT_nondt, .975))
)
kable(exp_effects, digits = 3, caption = "Experimental manipulation effects (validation).")
```

## Forest Plot

```{r forest, fig.width=10, fig.height=7}
all_eff <- rbind(
  data.frame(Parameter = covid_summary$Parameter,
             Category = "COVID", Mean = covid_summary$Mean,
             Lower = covid_summary$`Lower 95`, Upper = covid_summary$`Upper 95`),
  data.frame(Parameter = exp_effects$Effect,
             Category = "Experimental", Mean = exp_effects$Mean,
             Lower = exp_effects$Lower, Upper = exp_effects$Upper)
)
ggplot(all_eff, aes(x = Mean, y = reorder(Parameter, Mean), colour = Category)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = .25) +
  geom_vline(xintercept = 0, lty = 2, alpha = .5) +
  labs(title = "Posterior estimates with 95% credible intervals",
       x = "Effect size", y = NULL, colour = "Type") +
  theme_minimal() + theme(legend.position = "bottom")
```

## Group-Level Parameter Estimates

```{r group-params}
gp <- data.frame(
  Parameter = c("Drift Rate (v)", "Boundary (a)", "Non-decision Time (Ter)"),
  Mean = c(mean(samples$BUGSoutput$sims.list$drift_mean),
           mean(samples$BUGSoutput$sims.list$bound_mean),
           mean(samples$BUGSoutput$sims.list$nondt_mean)),
  SD = c(sd(samples$BUGSoutput$sims.list$drift_mean),
         sd(samples$BUGSoutput$sims.list$bound_mean),
         sd(samples$BUGSoutput$sims.list$nondt_mean)),
  `Lower 95` = c(quantile(samples$BUGSoutput$sims.list$drift_mean, .025),
                 quantile(samples$BUGSoutput$sims.list$bound_mean, .025),
                 quantile(samples$BUGSoutput$sims.list$nondt_mean, .025)),
  `Upper 95` = c(quantile(samples$BUGSoutput$sims.list$drift_mean, .975),
                 quantile(samples$BUGSoutput$sims.list$bound_mean, .975),
                 quantile(samples$BUGSoutput$sims.list$nondt_mean, .975)),
  check.names = FALSE
)
kable(gp, digits = 3, caption = "Group-level DDM parameter estimates.")
```

---

# Discussion

This study applied the EZBHDDM (Ch\'avez De la Pe\~na & Vandekerckhove, 2025) to investigate the cognitive effects of past COVID-19 exposure in a change-detection task. The hierarchical Bayesian approach decomposed performance into three cognitive parameters while accounting for uncertainty and individual differences.

The experimental manipulation effects serve as an internal validity check: Study Time affected drift rate (evidence accumulation), and Choice Urgency affected boundary separation (response caution), consistent with theoretical predictions. This confirms the model is capturing psychologically meaningful variation.

Regarding COVID-19 effects, the results should be interpreted in the context of the study's limitations. With 68 participants, the study has limited power to detect small effects. The COVID variable is based on self-report and the design is cross-sectional. Future research should employ longitudinal designs, serological confirmation, and multiple cognitive tasks to provide stronger evidence.

---

# References

Ch\'avez De la Pe\~na, A. F., & Vandekerckhove, J. (2025). An EZ Bayesian hierarchical drift diffusion model for response time and accuracy. *Psychonomic Bulletin \& Review, 32*(6), 3067--3087. https://doi.org/10.3758/s13423-025-02729-y

Hampshire, A., et al. (2021). Cognitive deficits in people who have recovered from COVID-19. *EClinicalMedicine, 39*, Article 101044. https://doi.org/10.1016/j.eclinm.2021.101044

Plummer, M. (2003). JAGS: A program for analysis of Bayesian graphical models using Gibbs sampling. In K. Hornik, F. Leisch, & A. Zeileis (Eds.), *Proceedings of the 3rd international workshop on distributed statistical computing*. Vienna, Austria.

Ratcliff, R. (1978). A theory of memory retrieval. *Psychological Review, 85*(2), 59--108. https://doi.org/10.1037/0033-295X.85.2.59

Wagenmakers, E.-J., van der Maas, H. L. J., & Grasman, R. P. P. P. (2007). An EZ-diffusion model for response time and accuracy. *Psychonomic Bulletin \& Review, 14*(1), 3--22. https://doi.org/10.3758/BF03194023
