---
title: "EZDDM JAGS Model with COVID indicator"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1: Set Up Working Environment

Prepare a fresh new R environment by removing previous variables and loading necessary libraries.

```{r}
# Fresh new working environment
graphics.off() # close windows
rm(list = ls())  # clear memory

# Load required libraries
library(readr)
library(R2jags)
library(dplyr)
```

## Step 2: Load and Prepare Data

We will load the data and clean it, selecting only the relevant variables for modeling.

```{r}
# Load and prepare the data
dfile <- "../data/processed/summarized_demographics_experiment_data_numeric.csv"
raw_data <- read_csv(dfile)

# Clean and prepare the data for DDM analysis
data_for_jags <- raw_data %>%
  select(mean_RT_sec, variance_RT_sec, total_accuracy, participant_id, covid, 
         nTrials, StudyTimeCoded, ChoiceUrgencyCoded, ProbOfChangeCoded,
         ProbeTypeCoded, age, raceNew_numeric, ethnicity_numeric,
         sex_numeric, education_numeric, version_numeric) %>%
  rename(meanRT  = mean_RT_sec, 
         varRT   = variance_RT_sec, 
         correct = total_accuracy) %>%
  na.omit()

data_for_jags$version_numeric <- 1 +
  1 * data_for_jags$StudyTimeCoded +
  2 * data_for_jags$ChoiceUrgencyCoded + 
  4 * data_for_jags$ProbOfChangeCoded + 
  8 * data_for_jags$ProbeTypeCoded
```

## Step 3: Modeling Overview

The next steps involve defining the JAGS model, setting up data, and running the model.
Below are pointers for these modeling tasks:

- Define the JAGS model, specifying prior distributions for hierarchical EZDDM parameters.
- Focus specifically on a predictive equation that either uses or predicts the covid indicator.
- Set up data for JAGS, including creating a list of variables for JAGS to use in the modeling.
- Specify initial values and run the JAGS model using `jags()` function.
- Extract and analyze the summary statistics for model parameters.

